---
title: "Global Suicide Rate Analysis Proposal"
author: "Qinyang Fang (qfang4),
  Siqi Gu (siqig2),
  Jinran Shi (jinrans2),
  Siyuan Teng (siyuant2) "
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    theme: cosmo
    toc: yes
  pdf_document: default
urlcolor: BrickRed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Nowadays, as the society develops at an increasingly rapid rate, more suicide incidents occur due to the increasing amount of stress and mental disorder illness including depression, anxiety disorders and so on. In order to investigate some feasible methods to prevent these incidents, the project aims to analyze the pattern of suicide rate in different countries, age group and gender.

## Data Background
  
  This suicide dataset was chosen from Kaggle and contains a total of 12 variables with 27,820 observations from 1985 to 2016. Variables include but not limited to country, year, sex, age and GDPs. For the country variable, most of the countries are in America and Europe. Genders are evenly distributed among this dataset which is 50% for each. A total of 6 age levels are used to indicate age. Below is the overall view of the original dataset.

```{r}
suicide = read.csv("suicide.csv")
head(suicide)
```

In this analysis, only data after year 2000 would be considered due to running time limit. Variables like suicides number and population would also be removed due to collinearity. Finally, the response variable, suicide rate, is planned to classify as three levels based on quantile. Therefore, the final dataset includes 7 variables.  

```{r,include = FALSE}
library(dplyr)
suicide = suicide[suicide$year >= 2000,]
suicide$year = as.factor(suicide$year)

#Drop variables
suicide = suicide[c(-5, -6,-8, -10, -12)]

#group levels
suicide$factor_rate = factor(case_when(
  suicide$suicides.100k.pop > quantile(suicide$suicides.100k.pop, 0.75) ~ "High",
  suicide$suicides.100k.pop > quantile(suicide$suicides.100k.pop, 0.25) ~ "Medium", 
  TRUE ~ "Low"
))

#Change Col Names
names(suicide) = c("country", "year", "sex", "age",  
                   "suicides_100k", "HDI_year", "gdp_capita", "factor_rate")

#replace HDI NAs
median = median(suicide[!is.na(suicide$HDI_year),]$HDI_year)
suicide$HDI_year[is.na(suicide$HDI_year)] = median
```

```{r}
head(suicide)
```

## Project Analysis Plan

In order to figure out the suicide rate of people around the world, both of the regression and classification models are planned to used in this analysis. At the end of the analysis, a "best" model will be reported for each type of models. For the regression part, models like linear, logistic and knn can be applied. On the other side, random forest, bagging and multinomial models are planned to predict classifications. 

## Expectations

An estimated 1.3 million adults attempt suicide each year, according to the Centers for Disease Control and Prevention (CDC). More than one in five people who died by suicide had expressed their suicide intent. 
Due to such a severe situation, the goal of this analysis is to prevent suicide. By predicting the suicide rates in specific age group or regions, an alarm can be sent to the local goverment and individuals. If more positive activities or necessary help could be provided, a lot of lives will be saved.
